---
title: "Train and test baseline models to predict MB subgroups"
author: "Steven Foltz"
date: "2022 - 2023"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
params:
  create_models: TRUE
  overwrite: FALSE
---

# Setup

```{r setup}

# get input parameters
create_models <- params$create_models
overwrite <- params$overwrite

# Source code and libraries
source(here::here("utils/color_schemes.R"))
source(here::here("utils/convert_gene_names.R"))
source(here::here("utils/modeling.R"))

# Define directories and input/output file paths
processed_data_dir <- here::here("processed_data")
models_dir <- here::here("models")
plots_dir <- here::here("plots")
plots_data_dir <- file.path(plots_dir, "data")

bulk_genex_filepath <- file.path(processed_data_dir, "bulk_genex.tsv")
bulk_metadata_filepath <- file.path(processed_data_dir, "bulk_metadata.tsv")

gene_map_filepath <- file.path(processed_data_dir, "gene_map.tsv")

baseline_filepath <- file.path(models_dir, "baseline.rds")

baseline_plot_df_filepath <- file.path(plots_data_dir, "baseline_plots_df.tsv")
baseline_kappa_plot_filepath <- file.path(plots_dir, "baseline_kappa.pdf")
baseline_accuracy_plot_filepath <- file.path(plots_dir, "baseline_accuracy.pdf")

# check that existing baseline model file will not be overwritten if overwrite is FALSE
if (create_models &
    (file.exists(baseline_filepath)) &
    !overwrite) {
  
  stop("Model output file already exists and overwrite is set to FALSE in analysis_notebooks/baseline.Rmd.")
  
}

# check that files exist if create_models is FALSE
if (!create_models & !file.exists(baseline_filepath)) {
  
  stop("Model output file does not exist and create_models is set to FALSE in analysis_notebooks/baseline.Rmd.")
  
}

# set subgroups analyzed in this notebook (canonical MB subgroups)
mb_subgroups <- c("G3", "G4", "SHH", "WNT")

# Read in essential data
bulk_genex_df <- readr::read_tsv(bulk_genex_filepath) |>
  tibble::column_to_rownames(var = "gene")

bulk_metadata_df <- readr::read_tsv(bulk_metadata_filepath) |>
  dplyr::filter(sample_accession %in% names(bulk_genex_df),
                subgroup %in% mb_subgroups)

bulk_genex_df <- bulk_genex_df |>
  dplyr::select(dplyr::all_of(bulk_metadata_df$sample_accession))

check_input_files(genex_df = bulk_genex_df,
                  metadata_df = bulk_metadata_df)

```

# Train and test MB subgroup models

### Create kTSP, RF, MM2S, and LASSO models

The following code creates both weighted and unweighted kTSP and RF models.
The MM2S and LASSO models use default settings.

Out of all the repeats, we select one 'official' repeat for visualization.
The 'official' repeat maximizes the median Kappa across all models within the repeat.

```{r}

if (create_models) {
  
  # kTSP and RF models with ktsp_weighted = TRUE and rf_weighted = TRUE (defaults)
  weighted_kTSP_RF_models_list <- run_many_models(genex_df = bulk_genex_df,
                                                  metadata_df = bulk_metadata_df,
                                                  labels = mb_subgroups,
                                                  model_types = c("ktsp", "rf"),
                                                  initial_seed = 44,
                                                  n_repeats = 10,
                                                  n_cores = 3,
                                                  ktsp_featureNo = 1000,
                                                  ktsp_n_rules_min = 5,
                                                  ktsp_n_rules_max = 50,
                                                  ktsp_weighted = TRUE,
                                                  rf_num.trees = 500,
                                                  rf_genes_altogether = 50,
                                                  rf_genes_one_vs_rest = 50,
                                                  rf_gene_repetition = 1,
                                                  rf_rules_altogether = 50,
                                                  rf_rules_one_vs_rest = 50,
                                                  rf_weighted = TRUE)
  
  # Add _weighted to names of kTSP and RF list elements
  weighted_kTSP_RF_models_list <- purrr::map(weighted_kTSP_RF_models_list,
                                             function(x) setNames(x,
                                                                  dplyr::recode(names(x),
                                                                                ktsp = "ktsp_weighted",
                                                                                rf = "rf_weighted")))
  
  # kTSP and RF models with ktsp_weighted = FALSE and rf_weighted = FALSE
  unweighted_kTSP_RF_models_list <- run_many_models(genex_df = bulk_genex_df,
                                                    metadata_df = bulk_metadata_df,
                                                    labels = mb_subgroups,
                                                    model_types = c("ktsp", "rf"),
                                                    initial_seed = 44,
                                                    n_repeats = 10,
                                                    n_cores = 3,
                                                    ktsp_featureNo = 1000,
                                                    ktsp_n_rules_min = 5,
                                                    ktsp_n_rules_max = 50,
                                                    ktsp_weighted = FALSE,
                                                    rf_num.trees = 500,
                                                    rf_genes_altogether = 50,
                                                    rf_genes_one_vs_rest = 50,
                                                    rf_gene_repetition = 1,
                                                    rf_rules_altogether = 50,
                                                    rf_rules_one_vs_rest = 50,
                                                    rf_weighted = FALSE)
  
  # Add _unweighted to names of kTSP and RF list elements
  unweighted_kTSP_RF_models_list <- purrr::map(unweighted_kTSP_RF_models_list,
                                               function(x) setNames(x,
                                                                    dplyr::recode(names(x),
                                                                                  ktsp = "ktsp_unweighted",
                                                                                  rf = "rf_unweighted")))
  
  # merge weighted and unweighted kTSP and RF model lists
  kTSP_RF_models_list <- purrr::map2(weighted_kTSP_RF_models_list,
                                     unweighted_kTSP_RF_models_list,
                                     function(x, y) c(x, y))
  
  # MM2S and LASSO models
  mm2s_lasso_models_list <- run_many_models(genex_df = bulk_genex_df,
                                            metadata_df = bulk_metadata_df,
                                            labels = mb_subgroups,
                                            model_types = c("mm2s", "lasso"),
                                            initial_seed = 44,
                                            n_repeats = 10,
                                            n_cores = 3,
                                            mm2s_gene_map_filepath = gene_map_filepath)
  
  # merge kTSP, RF, MM2S, and LASSO model lists
  baseline_list <- purrr::map2(kTSP_RF_models_list,
                               mm2s_lasso_models_list,
                               function(x, y) c(x, y)) |>
    purrr::map(function(x) x[!duplicated(names(x))]) # remove duplicate list items
  
  # select "official" repeat with greatest median Kappa across models
  model_types <- c("ktsp_weighted", "ktsp_unweighted",
                   "rf_weighted", "rf_unweighted",
                   "lasso", "mm2s")
  
  official_model <- which.max(unlist(purrr::map(baseline_list,
                                                function(x) median(unlist(purrr::map(model_types,
                                                                                     function(y) x[[y]]$cm$overall[["Kappa"]]))))))
  
  for (model_index in 1:length(baseline_list)) {
    
    if (model_index == official_model) {
      
      baseline_list[[model_index]][["official_model"]] <- TRUE
      
    } else {
      
      baseline_list[[model_index]][["official_model"]] <- FALSE
      
    }
    
  }
  
  # write models to file
  readr::write_rds(x = baseline_list,
                   file = baseline_filepath)
  
} else { # if create_models = FALSE and models already exist
  
  baseline_list <- readr::read_rds(baseline_filepath)
  
}

```

# Visualize kappa and accuracy results

### Gather data

```{r}

plot_df <- purrr::map(baseline_list,
                      function(x) return_model_metrics(single_repeat = x,
                                                       model_types = c("ktsp_weighted",
                                                                       "ktsp_unweighted",
                                                                       "rf_weighted",
                                                                       "rf_unweighted",
                                                                       "mm2s",
                                                                       "lasso"),
                                                       metadata_df = bulk_metadata_df,
                                                       labels = c("G3",
                                                                  "G4",
                                                                  "WNT",
                                                                  "SHH"),
                                                       platforms = c("Array",
                                                                     "RNA-seq"))) |>
  reshape2::melt(id.vars = c("model_type",
                             "platform",
                             "study",
                             "which_studies",
                             "official_model",
                             "total_samples",
                             "subgroup",
                             "subgroup_samples",
                             "metric",
                             "value")) |>
  dplyr::rename(repeat_index = L1)

readr::write_tsv(x = plot_df,
                 file = baseline_plot_df_filepath)

```

### Plot Kappa

```{r}

set.seed(1)

baseline_kappa_plot <- plot_df |>
  dplyr::filter(metric == "Kappa") |>
  ggplot2::ggplot(ggplot2::aes(x = model_type,
                               y = value,
                               fill = platform,
                               color = platform)) +
  ggplot2::geom_violin(position = ggplot2::position_dodge(0.7),
                       alpha = 0.5,
                       color = NA) +
  ggplot2::geom_point(shape = 16,
                      position = ggplot2::position_jitterdodge(jitter.width = 0.1,
                                                               jitter.height = 0)) +
  ggplot2::labs(x = "Method",
                y = "Kappa") +
  ggplot2::theme_bw()

baseline_kappa_plot

ggplot2::ggsave(filename = baseline_kappa_plot_filepath,
                plot = baseline_kappa_plot,
                height = 7,
                width = 7)

```

### Plot accuracy

```{r}

baseline_accuracy_plot <- plot_df |>
  dplyr::filter(metric == "Balanced Accuracy") |>
  ggplot2::ggplot(ggplot2::aes(x = model_type,
                               y = value,
                               fill = platform,
                               color = subgroup)) +
  ggplot2::geom_violin(draw_quantiles = 0.5,
                       show.legend = FALSE,
                       position = ggplot2::position_dodge(0.7)) +
  ggplot2::geom_point(shape = 16,
                      position = ggplot2::position_jitterdodge(jitter.width = 0.1,
                                                               jitter.height = 0),
                      show.legend = FALSE) +
  ggplot2::facet_wrap(~ subgroup,
                      ncol = 2) +
  ggplot2::scale_color_manual(values = okabe_ito_colors) +
  ggplot2::scale_fill_manual(values = c("#FFFFFF", "#FFFFFF")) + 
  ggplot2::labs(x = "Subgroup",
                y = "Balanced Accuracy") +
  ggplot2::theme_bw()

baseline_accuracy_plot

ggplot2::ggsave(filename = baseline_accuracy_plot_filepath,
                plot = baseline_accuracy_plot,
                height = 7,
                width = 7)

```

# Session info

```{r session_info}
sessionInfo()
```