---
title: "Labeling non-malignant cells in the Hovestadt et al. Smart-seq2 dataset"
author: "Jaclyn Taroni"
date: '2025'
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
---

We do not have author-derived cell type labels for the [Hovestadt, et al. 2019.](https://www.nature.com/articles/s41586-019-1434-6) Smart-seq2 dataset.
Based on [Extended Data Figure 3](https://pmc.ncbi.nlm.nih.gov/articles/PMC6754173/figure/F9/), we expect a small number of immune cells and oligodendrocytes to be present in the dataset, with the overwhelming majority of cells being malignant cells.
To address the question of labeling non-malignant cells, we took two approaches in `scripts/classify_singler_aucell.R`:

- Used SingleR ([Aran et al. 2019.](https://doi.org/10.1038/s41590-018-0276-y)) with [the `BlueprintEncodeData`](https://bioconductor.org/packages/release/data/experiment/vignettes/celldex/inst/doc/userguide.html#32_BlueprintENCODE) as a reference with the goal of identifying immune cells.
- Used AUCell ([Aibar et al. 2017.](https://doi.org/10.1038/nmeth.4463)) with PanglaoDB ([Franzén et al. 2019.](https://academic.oup.com/database/article/doi/10.1093/database/baz046/5427041)) marker genes for oligodendrocytes with the goal of identifying oligodendrocytes. 
We use AUCell's automated thresholding to make oligodendrocyte assignments.

You can see how the references were prepared in `scripts/set_up_annotation_references.R`.

A lot of this code was adapted from Childhood Cancer Data Lab training material on merging `SingleCellExperiment` objects.

## Set up

### Libraries

```{r}
suppressMessages(library(SingleCellExperiment))
library(patchwork)
```

### Color scheme

```{r}
source(here::here("utils/color_schemes.R"))
```

### Functions

```{r}
# Adapted from https://github.com/AlexsLemonade/training-modules/blob/32c765735b1e3a9c9df390d12af362a74f527d8f/scRNA-seq-advanced/02-dataset_integration.Rmd#L358-L412
#' Custom function to format an SCE before merging
#'
#' @param sce SCE object to format
#' @param sample_name Name of the sample
#' @param shared_genes Vector of shared genes across all SCE objects
#'
#' @returns An updated SCE object ready for merging
format_sce <- function(
  sce, 
  sample_name, 
  shared_genes
) {
  
  ### Remove the single-sample reduced dimensions 
  # We do this first since it makes the object a lot smaller for the rest of this code!
  reducedDims(sce) <- NULL
  
  ### Add dedicated sample indicator column to the colData slot
  # Recall, the `sce$` shortcut points to the colData
  sce$sample <- sample_name

  ### Ensure objects have the same genes in the same order
  # Use the shared_genes vector to index genes to the intersection
  # Doing this both subsets to just those genes, and reorders!
  sce <- sce[shared_genes, ]

  ### Ensure cell ids are identifiable and fully unique 
  # Update the SCE object column names (cell ids) by prepending the `sample_name`
  colnames(sce) <- glue::glue("{sample_name}-{colnames(sce)}")
  
  ### Finally, we can return the formatted SCE object
  return(sce)  
}
```

### Files and directories

```{r}
# Directories
processed_data_dir <- here::here("processed_data")
single_cell_data_dir <- here::here(processed_data_dir, "single_cell")
smartseq_data_dir <- here::here(single_cell_data_dir, "GSE119926/sce")

# Grab all the SingleCellExperiment files
sce_files <- fs::dir_ls(path = smartseq_data_dir,
                        glob = "*_sce.rds")

# Extract sample titles from the file names
sce_sample_titles <- stringr::word(names(sce_files), start = -1, sep = "/") |>
  stringr::str_remove_all("\\_sce.rds")

# Fail if there are any sample title collisions
if (any(duplicated(sce_sample_titles))) {

  stop("Duplicate single-cell sample title detected!")

} else {

  # Name the vector using the sample title
  names(sce_files) <- sce_sample_titles

}

# Pseudobulk metadata file
pseudobulk_meta_file <- fs::path(
  single_cell_data_dir, 
  "pseudobulk_metadata.tsv"
)
```

## Read in and process data

Read in metadata.

```{r}
metadata_df <- readr::read_tsv(pseudobulk_meta_file)
```

Read in `SingleCellExperiment` objects and format for merging.

```{r}
# Read in SingleCellExperiment objects into a list
sce_list <- sce_files |>
  purrr::map(readr::read_rds)

# Get all the shared genes
shared_genes <- sce_list |>
  # get rownames (genes) for each SCE in sce_list
  purrr::map(rownames) |>
  # reduce to the _intersection_ among lists
  purrr::reduce(intersect)

# Process all the SingleCellExperiment in the list to prepare for merging
sce_list_formatted <- purrr::map2(
  sce_list,
  sce_sample_titles,
  \(sce, sample_name) format_sce(sce, sample_name, shared_genes) 
)
```

### Merge objects and log2-transform the TPM

```{r}
# Merge (not integrated!!!)
merged_sce <- do.call(cbind, sce_list_formatted)

# Log2 transform TPM to populate logcounts slot -- this is otherwise 
# scater::logNormCounts applied to TPM because of the processing that happens in
# 02-gather-data.R
logcounts(merged_sce) <- log2(counts(merged_sce) + 1)
```

Remove large objects from the environment because we no longer need them.

```{r}
# Remove sce_list and sce_list_formatted
rm(sce_list)
rm(sce_list_formatted)
```

## Dimensionality reduction

### HVGs and PCA

```{r}
# Adapted from https://github.com/AlexsLemonade/training-modules/blob/32c765735b1e3a9c9df390d12af362a74f527d8f/scRNA-seq-advanced/02-dataset_integration.Rmd
# Specify the number of high variance genes to use
num_genes <- 2000

# Calculate variation for each gene
gene_variance <- scran::modelGeneVar(merged_sce,
                                     # specify the grouping column:
                                     block = merged_sce$sample)

# Get the top `num_genes` high-variance genes to use for dimension reduction
hv_genes <- scran::getTopHVGs(gene_variance,
                              n = num_genes)

# Use batchelor to calculate PCA for merged_sce, considering only
#  the high-variance genes
# We'll need to include the argument `preserve.single = TRUE` to get
#  a single matrix with all samples and not separate matrices for each sample
merged_pca <- batchelor::multiBatchPCA(merged_sce,
                                       subset.row = hv_genes,
                                       batch = merged_sce$sample,
                                       preserve.single = TRUE)
```

### UMAP

```{r}
# Adapted from https://github.com/AlexsLemonade/training-modules/blob/32c765735b1e3a9c9df390d12af362a74f527d8f/scRNA-seq-advanced/02-dataset_integration.Rmd
# Add PCA results to merged object
reducedDim(merged_sce, "PCA") <- merged_pca[[1]]

# Run UMAP
merged_sce <- scater::runUMAP(merged_sce)
```

#### Plotting

To facilitate plotting, we're going to add the true subgroup labels to the `colData`.

```{r}
# Extract colData and wrangle
coldata_df <- colData(merged_sce) |>
  as.data.frame() |>
  tibble::rownames_to_column("id") |>
  dplyr::mutate(title = stringr::str_extract(id, "[^-]+"))

# Add in relevant metadata from pseudobulk metadata
coldata_df <- coldata_df |>
  dplyr::left_join(
    dplyr::select(metadata_df,
                  title,
                  is_PDX,
                  subgroup,
                  subtype),
    by = "title"
  )

# Add back in
colData(merged_sce) <- DataFrame(
  coldata_df,
  row.names = rownames(coldata_df)
)
```

Plot a UMAP colored by true subgroup labels.

```{r}
# UMAPs scaled together when calculated from the merged SCE
subgroup_umap <- scater::plotUMAP(
  merged_sce,
  color_by = "subgroup",
  # Some styling to help us see the points:
  point_size = 0.5,
  point_alpha = 0.2
) +
  ggplot2::scale_color_manual(values = subgroup_colors) + 
  ggplot2::guides(color = ggplot2::guide_legend(
    override.aes = list(size = 3, 
                        alpha = 1))
  ) +
  ggplot2::ggtitle("Subgroup")
```

Plot UMAP coloring cells based on whether or not something got called an oligodendrocyte using AUCell methodology.

```{r}
oligodendrocyte_umap <- scater::plotUMAP(
  merged_sce,
  color_by = "oligodendrocyte_assignment",
  # Some styling to help us see the points:
  point_size = 0.5,
  point_alpha = 0.2
) +
  ggplot2::guides(color = ggplot2::guide_legend(
    override.aes = list(size = 3, 
                        alpha = 1))
  ) +
  ggplot2::ggtitle("Oligodendrocyte Assignment")
```

Make a bar plot counting the number of SingleR annotations.

```{r}
singler_barplot <- coldata_df |> 
  ggplot2::ggplot(
    ggplot2::aes(
      x = pruned.labels
    )
  ) + 
  ggplot2::geom_bar(stat = "count") +
  ggplot2::theme_bw() +
  ggplot2::theme(axis.text.x = ggplot2::element_text(
      angle = 90, vjust = 0.5, hjust = 1
    )
  ) +
  ggplot2::ggtitle("SingleR Assignment") +
  ggplot2::xlab("Cell Type Label")
```

Compile figure.

```{r}
(subgroup_umap + oligodendrocyte_umap) / singler_barplot
```

## Takeaways

We do not see any immune cells in the SingleR annotations.
We also don't see a coherent group of oligodendrocytes separate from the tumor cells.
In reviewing these results, I thought it might be possible that non-malignant cells were removed from the data on GEO.
If we look at the description a sample on GEO, it typically says "N high quality malignant cells." 
Here's an example: <https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3905423>
Therefore, I think it's safe to assume that all cells in this dataset are malignant cells.

## Session Info

```{r}
sessionInfo()
```
