---
title: "Explore model accuracy using a targeted gene set"
author: "Steven Foltz"
date: "2025"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
---

# Rationale

Experiments that generate RNA expression data may use a targeted gene panel as a more cost-effective approach to analyze a subset of genes thought to have greater relevance for disease, especially in clinical settings.
However, when a model is trained using whole transcriptome data, the genes selected for that model will generally show only partial or no overlap with a targeted gene panel.
Even though kTSP and RF models can tolerate missing genes, prediction performance on new samples will likely suffer as a result of having fewer gene:gene comparisons to score.
A model trained using a targeted panel of genes from the start could be applied to sample data that includes more genes, however there may be a decrease in model performance due to starting with a smaller feature space.
This leads to two key questions:

1. Given a targeted gene set of cancer-related genes, how does model performance compare between targeted gene set models and full gene set models?
2. Does the identity of the genes in the targeted gene set matter? In other words, can a random gene set of the same size perform just as well as the targeted gene set?

We will use the [Nanostring nCounter PanCancer IO 360 Panel](https://nanostring.com/products/ncounter-assays-panels/oncology/pancancer-io-360/) targeted gene set published by Nanostring, available for download [here](https://nanostring.com/products/ncounter-assays-panels/panel-selection-tool/) and found in this repository at `processed_data/NS_IO_360_v1.0_Genes.tsv`.

# Setup

```{r setup}

# Source code and libraries
source(here::here("utils/color_schemes.R"))
source(here::here("utils/modeling.R"))
source(here::here("utils/convert_gene_names.R"))
library(patchwork)

# Define directories and input/output file paths
processed_data_dir <- here::here("processed_data")
models_dir <- here::here("models")
plots_dir <- here::here("plots")
plots_data_dir <- here::here(plots_dir, "data")

bulk_genex_filepath <- here::here(processed_data_dir, "bulk_genex.tsv")
bulk_metadata_filepath <- here::here(processed_data_dir, "bulk_metadata.tsv")

baseline_filepath <- here::here(models_dir, "baseline.rds")
targeted_filepath <- here::here(models_dir, "targeted_gene_set.rds")
random_filepath <- here::here(models_dir, "random_gene_sets.rds")

baseline_plot_df_filepath <- here::here(plots_data_dir, "baseline_plots_df.tsv")
targeted_plot_df_filepath <- here::here(plots_data_dir, "targeted_plot_df.tsv")
random_plot_df_filepath <- here::here(plots_data_dir, "random_plot_df.tsv")

combined_plot_filepath <- here::here(plots_dir, "targeted_random_full_kappa_accuracy.pdf")

# set subgroups analyzed in this notebook (canonical MB subgroups)
mb_subgroups <- c("G3", "G4", "SHH", "WNT")

# Read in essential data
bulk_genex_df <- readr::read_tsv(bulk_genex_filepath,
                                 show_col_types = FALSE) |>
  tibble::column_to_rownames(var = "gene")

bulk_metadata_df <- readr::read_tsv(bulk_metadata_filepath,
                                    show_col_types = FALSE) |>
  dplyr::filter(sample_accession %in% names(bulk_genex_df),
                subgroup %in% mb_subgroups)

bulk_genex_df <- bulk_genex_df |>
  dplyr::select(dplyr::all_of(bulk_metadata_df$sample_accession))

check_input_files(genex_df = bulk_genex_df,
                  metadata_df = bulk_metadata_df)


```

# Visualize Kappa and accuracy results

### Gather targeted, random, and baseline data

```{r}

targeted_list <- readr::read_rds(targeted_filepath)
random_list <- readr::read_rds(random_filepath)

model_types <- c("ktsp", "rf", "lasso")

# targeted model metrics
targeted_plot_df <- targeted_list |>
  purrr::map(return_model_metrics,
             model_types = model_types,
             metadata_df = bulk_metadata_df,
             labels = mb_subgroups,
             platforms = c("Array", "RNA-seq")) |>
  purrr::list_rbind(names_to = "repeat_index") |>
  dplyr::mutate(model_type = dplyr::case_match(model_type,
                                               "ktsp" ~ "kTSP (unw)",
                                               "rf" ~ "RF (w)",
                                               "lasso" ~ "LASSO"))

readr::write_tsv(x = targeted_plot_df,
                 file = targeted_plot_df_filepath)

# random model metrics
random_plot_df <- random_list |>
  purrr::map(\(x) x |>
               purrr::map(return_model_metrics,
                          model_types = model_types,
                          metadata_df = bulk_metadata_df,
                          labels = mb_subgroups,
                          platforms = c("Array", "RNA-seq")) |>
               purrr::list_rbind(names_to = "repeat_index") |>
               dplyr::mutate(model_type = dplyr::case_match(model_type,
                                                            "ktsp" ~ "kTSP (unw)",
                                                            "rf" ~ "RF (w)",
                                                            "lasso" ~ "LASSO"))) |>
  purrr::list_rbind(names_to = "random_index")

readr::write_tsv(x = random_plot_df,
                 file = random_plot_df_filepath)

# baseline model metrics
baseline_plot_df <- readr::read_tsv(file = baseline_plot_df_filepath,
                                    show_col_types = FALSE) |>
  dplyr::filter(model_type %in% c("kTSP (unw)", "RF (w)", "LASSO"))

# combined dfs

compare_metrics_plot_df <- dplyr::bind_rows(targeted_plot_df |>
                                              dplyr::mutate(gene_set = "Targeted"),
                                            random_plot_df |>
                                              dplyr::mutate(gene_set = "Random"),
                                            baseline_plot_df |>
                                              dplyr::mutate(gene_set = "Full"))

```

# Compare kappa values

### Is there a statistical difference?

```{r}

test_list <- purrr::map(c("kTSP (unw)", "RF (w)", "LASSO"),
                        \(mt) purrr::map(c("Array", "RNA-seq"),
                                         \(p) {
                                           
                                           wilcox.test(x = targeted_plot_df |>
                                                         dplyr::filter(model_type == mt,
                                                                       platform == p) |> 
                                                         dplyr::pull(value),
                                                       y = baseline_plot_df |> 
                                                         dplyr::filter(model_type == mt, 
                                                                       platform == p) |> 
                                                         dplyr::pull(value))
                                           
                                         }
                        ) |> setNames(c("Array", "RNA-seq"))
                        
) |> setNames(c("kTSP (unw)", "RF (w)", "LASSO"))


```

### Visualize kappa difference 
```{r}

kappa_plot_object <- compare_metrics_plot_df |>
  dplyr::filter(metric == "Kappa") |>
  ggplot2::ggplot(ggplot2::aes(x = platform,
                               y = value,
                               color = gene_set)) +
  ggplot2::geom_boxplot() +
  ggplot2::facet_wrap(~ model_type,
                      nrow = 1) +
  ggplot2::labs(x = "Method",
                y = "Kappa",
                color = "Gene Set",
                title = "Kappa comparison across gene sets") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom",
                 legend.direction = "horizontal")

accuracy_plot_object <- compare_metrics_plot_df |>
  dplyr::filter(metric == "Balanced Accuracy") |>
  ggplot2::ggplot(ggplot2::aes(x = platform,
                               y = value,
                               color = gene_set)) +
  ggplot2::geom_boxplot() +
  ggplot2::facet_grid(rows = dplyr::vars(subgroup), cols = dplyr::vars(model_type)) +
  ggplot2::labs(x = "Method",
                y = "Balanced Accuracy",
                color = "Gene Set",
                title = "Balanced accuracy comparison across gene sets") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom",
                 legend.direction = "horizontal")

combined_plot <- kappa_plot_object / accuracy_plot_object +
  plot_annotation(tag_levels = "a") +
  plot_layout(guides = "collect", heights = c(1,4)) &
  ggplot2::theme(legend.position = "bottom",
                 legend.direction = "horizontal")

ggplot2::ggsave(filename = combined_plot_filepath,
                plot = combined_plot,
                height = 7,
                width = 7)

combined_plot

```


# Session info

```{r session_info}
sessionInfo()
```
