---
title: "GSE124814 metadata"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r file paths}
data_dir <- here::here("data")
GSE124814_folder <- file.path(data_dir, "GSE124814")

GSE124814_metadata_input_filename <- file.path(GSE124814_folder,
                                               "GSE124814_sample_descriptions.xlsx")
GSE124814_metadata_output_filename <- file.path(GSE124814_folder,
                                                "GSE124814_metadata.tsv")
GSE124814_duplicate_samples_filename <- file.path(GSE124814_folder,
                                                  "GSE124814_duplicate_samples.tsv")
GSE124814_experiment_accessions_output_filename <- file.path(data_dir,
                                                             "GSE124814_experiment_accessions.tsv")

```

```{r load libraries}
library(tidyverse)
```

[Weishaupt et al.](https://academic.oup.com/bioinformatics/article/35/18/3357/5305636) did a lot of heavy lifting to collect publicly-available MB samples, including harmonizing metadata:

- mapped the original subgroups (`subgroup_supplied_original`) to consensus subgroups (`subgroup_supplied_renamed`)
- identified duplicate samples (merged duplicate samples together by averaging expression values)
- limited to just MB and normal cerebellum tissues

Citation: Weishaupt, H. et al. Batch-normalization of cerebellar and medulloblastoma gene expression datasets utilizing empirically defined negative control genes. Bioinformatics 35, 3357–3364 (2019)

```{r}
GSE124814_samples_df_column_names <- c("Sample_name",
                                       "title",
                                       "CEL_file",
                                       "source_name",
                                       "organism",
                                       "molecule",
                                       "label",
                                       "chip_name",
                                       "age",
                                       "sex", # was gender
                                       "histology",
                                       "brain_region",
                                       "death",
                                       "metastatic_stage",
                                       "follow_up",
                                       "subgroup_supplied_original",
                                       "subgroup_supplied_renamed",
                                       "subgroup_relabeled",
                                       "description",
                                       "description2")

# when reading in the xlsx file,
# 1. use the column names above
# 2. skip final column
# 3. ignore first two rows
GSE124814_samples_df <- readxl::read_xlsx(GSE124814_metadata_input_filename,
                                          col_names = GSE124814_samples_df_column_names,
                                          col_types = "text",
                                          skip = 2) %>%
  separate(title, # separate title into experiment accession and sample id
           into = c("experiment_accession", "id"),
           sep = "-",
           extra = "merge") %>% # split on the first hyphen and keep rest intact
  # reformat E-MTAB-292 experiment accession name
  mutate(experiment_accession = str_replace(experiment_accession, "EMTAB", "E-MTAB-")) %>%
  # GSE124814 merged duplicate samples together by averaging expression values
  mutate(is_duplicate = str_detect(description, "average")) %>%
  # "NA" subgroup corresponds to "Normal"
  mutate(subgroup_supplied_renamed = case_when(subgroup_supplied_renamed == "NA" ~ "Normal",
                                               TRUE ~ subgroup_supplied_renamed)) %>%
  # treat "Unknown" subgroup as NA -- we need to look further into these
  mutate(subgroup_supplied_renamed = na_if(x = subgroup_supplied_renamed,
                                           y = "Unknown")) %>%
  mutate(sex = na_if(x = sex, # can potentially re-label these based on chrY expression (total 195)
                     y = "NA"), # or cross-reference with refine.bio metadata
         sex = case_when(sex %in% c("f", "female") ~ "F", # use F as default
                         sex %in% c("m", "male") ~ "M", # use M as default
                         TRUE ~ sex)) %>%
  # isolate the sample accession as its own column ("reanalysis of SAMPLE_ACCESSION ...")
  rowwise() %>%
  mutate(sample_accession = str_split(description,
                                      pattern = " ",
                                      simplify = TRUE)[3]) %>%
  ungroup() %>%
  # need to consider "age" -- lots of variation in formats, check with refine.bio
  # deselect columns that are not important to us and/or have only one value
  select(-c("CEL_file", # "Reanalysis of existing record"
            "organism", # "Homo sapiens"
            "molecule", # variations of "total RNA"
            "label", # variations of "biotin"
            "description2")) %>% # repeat of Sample_name
  write_tsv(file = GSE124814_metadata_output_filename)

```

### GSE124814 experiment accessions (total samples from each relabeled subgroup)

```{r GSE124814 accessions, results = "asis"}
GSE124814_samples_df %>%
  count(experiment_accession, subgroup_supplied_renamed) %>%
  pivot_wider(names_from = subgroup_supplied_renamed,
              values_from = n) %>%
  left_join(GSE124814_samples_df %>%
              group_by(experiment_accession) %>%
              summarize("duplicates" = sum(is_duplicate)),
            by = "experiment_accession") %>%
  left_join(GSE124814_samples_df %>%
              count(experiment_accession) %>%
              rename("Total" = "n"),
            by = "experiment_accession") %>%
  replace(is.na(.), 0) %>%
  arrange(desc(Total)) %>%
  # add column sums, thanks to https://stackoverflow.com/a/50322272
  bind_rows(summarise_all(., ~if(is.numeric(.)) sum(.) else "Total")) %>%
  knitr::kable()

GSE124814_samples_df %>%
  select(experiment_accession) %>%
  unique() %>%
  write_tsv(file = GSE124814_experiment_accessions_output_filename,
            col_names = FALSE)
```
### Mapping original subgroups to consensus subgroups

Mapping was done by Weishaupt et al.
We have further updated the mapping with the following:

- `NA` is an unknown subgroup ("Unknown" in the Weishaupt metadata)
- `Normal` was given as "NA" in the Weishaupt metadata

```{r subgroup mapping, results = "asis"}
# Display table of original subgroups mapped to consensus subgroups  
GSE124814_samples_df %>%
  select(subgroup_supplied_original,
         subgroup_supplied_renamed) %>%
  table(useNA = "ifany") %>%
  knitr::kable()
```


### Duplicated patients identified by Weishaupt et al.

Note: could be 2+ samples per patient

```{r duplicate samples, results = "asis"}
GSE124814_samples_df %>%
  filter(is_duplicate) %>%
  count(experiment_accession, subgroup_supplied_renamed) %>%
  pivot_wider(names_from = subgroup_supplied_renamed,
              values_from = n) %>%
  replace(is.na(.), 0) %>%
  knitr::kable()

# save duplicate samples to file
GSE124814_samples_df %>%
  filter(is_duplicate) %>%
  select(experiment_accession, sample_accession) %>%
  write_tsv(file = GSE124814_duplicate_samples_filename) 
```