---
title: "Inspecting genes included in baseline models"
author: "Jaclyn Taroni"
date: "2025"
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
---

## Set up

### Directories

```{r}
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
models_dir <- file.path(root_dir, "models")
results_dir <- file.path(root_dir, "results")
dir.create(results_dir, showWarnings = FALSE)
```

### Input

```{r}
baseline_models_filepath <- file.path(models_dir, "baseline.rds")
```

### Output

```{r}
genes_list_filepath <- file.path(results_dir, 
                                 "genes_in_kTSP_unw_and_RF_w_official_models.tsv")
```

## Get pair-based models

We are going to look at the weighted version of the kTSP and RF models in the "official" repeat (i.e., the repeat selected by maximizing Kappa).

```{r}
baseline_models <- readr::read_rds(baseline_models_filepath)
official_model <- which(purrr::map_lgl(baseline_models, 
                                           \(x) x[["official_model"]]))
ktsp_model <- baseline_models[[official_model]]$ktsp_unweighted$classifier
rf_model <- baseline_models[[official_model]]$rf_weighted$classifier
```

## Extract genes used in models

### kTSP

kTSP models report the gene pairs used in the rules for each class (subgroup).
We'll want to keep track of the subgroup, but get this in a longer format.

```{r}
tsp_df <- ktsp_model$classifiers |>
  purrr::map(\(x)
             data.frame(x$TSP)) |>
  dplyr::bind_rows(.id = "subgroup") |>
  tidyr::pivot_longer(cols = dplyr::starts_with("gene"),
                      names_to = "gene_in_pair",
                      values_to = "gene") |>
  dplyr::mutate(model = "kTSP") |>
  dplyr::rename(gene_subgroup = subgroup)
```

### RF

The RF model just reports the genes that are included.

```{r}
rf_df <- data.frame(gene = rf_model$RF_scheme$genes) |>
  dplyr::mutate(model = "RF")
```

### Bind and save

```{r}
tsp_df |>
  dplyr::bind_rows(rf_df) |>
  readr::write_tsv(genes_list_filepath)
```

### What genes are overlapping between the two models?

Create a list to be passed to the UpSetR functions.

```{r}
genesets_list <- unique(tsp_df$gene_subgroup) |>
  purrr::set_names(~ paste0("kTSP_", .)) |>
  purrr::map(
    \(x) tsp_df |>
      dplyr::filter(gene_subgroup == x) |>
      dplyr::pull(gene)
    )
genesets_list$RF <- rf_df$gene
```

#### UpSet Plot

```{r}
UpSetR::upset(UpSetR::fromList(genesets_list))
```

For each subgroup's rules included in the kTSP model, some of those genes overlap with the RF model genes.
However, most of the genes included in the RF model aren't represented in the kTSP rules.

## Session Info

```{r}
sessionInfo()
```
